---
title: "Data & provenance"
vignette: >
  %\VignetteIndexEntry{Data & provenance}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(cfbstats)
library(dplyr)
library(reactable)
library(targets)
root <- Sys.getenv("CFBSTATS_ROOT")
if (nzchar(root)) knitr::opts_knit$set(root.dir = root)
```

This project is built to be **faithful to the raw data and transparent about
every transformation**. Two things support that: a **data dictionary** for each
source, and a **per-step audit log** generated by the pipeline (decision 0005).

## Sources

We build on the [College Football Data](https://collegefootballdata.com) API,
plus [nflverse](https://nflverse.nflreadr.com/) for post-draft NFL outcomes
(rosters, statistics, and career summaries) linked to draft picks by draft slot
(decision 0014). Each dataset has a dictionary (shipped in the package under
`inst/dict/`) describing its grain, key, and columns.

```{r}
#| label: dictionaries
dict_dir <- system.file("dict", package = "cfbstats")
dict_files <- list.files(dict_dir, pattern = "\\.yml$", full.names = TRUE)
purrr::map_dfr(dict_files, function(f) {
  d <- yaml::read_yaml(f)
  tibble::tibble(
    dataset = d$dataset, source = d$source, grain = d$grain,
    key = d$key, n_columns = length(d$columns)
  )
}) |>
  reactable(searchable = TRUE, striped = TRUE, defaultPageSize = 10)
```

### Column-level detail

```{r}
#| label: dictionary-columns
purrr::map_dfr(dict_files, function(f) {
  d <- yaml::read_yaml(f)
  purrr::map_dfr(d$columns, function(col) {
    tibble::tibble(
      dataset = d$dataset, column = col$name, type = col$type,
      provenance = col$provenance %||% NA_character_,
      description = col$description
    )
  })
}) |>
  reactable(
    searchable = TRUE, filterable = TRUE, striped = TRUE,
    groupBy = "dataset", defaultPageSize = 15
  )
```

## Provenance / audit log

Every tracked pipeline step records a uniform row: input/output row and column
counts, rows dropped, key coverage, NA cells, a schema hash, and the governing
data contract's pass/fail (decision 0005). This is where **silent join loss and
schema drift become visible**.

```{r}
#| label: audit-log
tar_read(audit_log) |>
  select(step, stage, n_rows_in, n_rows_out, rows_delta,
         n_cols_out, n_keys_out, n_na_cells, contract_passed, schema_hash) |>
  reactable(
    striped = TRUE, defaultPageSize = 12,
    columns = list(
      rows_delta = colDef(
        style = function(value) {
          if (is.na(value)) return(NULL)
          if (value < 0) list(color = "#b00020") else if (value > 0)
            list(color = "#0059b3") else NULL
        }
      ),
      contract_passed = colDef(name = "contract")
    )
  )
```

::: {.callout-note}
## Reading the log
`build_player_season` shows the large negative `rows_delta` where the **long**
stats collapse to one identity row per player-season. `link_coaches` shows a
positive delta: some schools carry more than one head coach in a season
(midseason changes), which fans the join out — exactly the kind of row change
we want surfaced rather than hidden.
:::
