---
title: "Explore"
vignette: >
  %\VignetteIndexEntry{Explore}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(cfbstats)
library(dplyr)
library(targets)
root <- Sys.getenv("CFBSTATS_ROOT")
if (nzchar(root)) {
  knitr::opts_knit$set(root.dir = root)
}
```

Interactive exemplars for the three chart families the project uses: **time
series**, **2D**, and **3D**. Each reads a pipeline target, so the input → chart
lineage is explicit. The default stack is R htmlwidgets (`plotly`, `ggiraph`);
Observable JS is reserved for bespoke explorables (a later addition).

```{r}
#| label: data
picks <- tar_read(picks) |>
  filter(year >= 2013, !is.na(height), !is.na(weight), !is.na(overall))
```

## Time series — drafts per year

```{r}
#| label: fig-timeseries
viz_drafts_by_season(tar_read(model_table))
```

## 2D — draft capital vs. physicals (hover for player)

Each point is a drafted player; hover for the name. Color encodes draft round.

```{r}
#| label: fig-2d
viz_scatter_2d(
  dplyr::slice_sample(picks, n = min(1500, nrow(picks))),
  x = weight,
  y = height,
  tooltip = name,
  color = round
)
```

## 3D — height × weight × overall pick

```{r}
#| label: fig-3d
viz_scatter_3d(
  picks,
  x = "height",
  y = "weight",
  z = "overall",
  color = "round"
)
```

## NFL outcomes — how long drafted players last

Each point is a drafted player bridged to nflverse (decision 0014), placed by
draft round against the number of distinct **seasons they spent on an NFL
roster** — the project's primary longevity measure. The **#1 overall picks** are
highlighted against the rest of the field; the grey bar marks each round's
median. Hover for the player. Recent draft classes are still active, so their
counts are **right-censored** lower bounds.

```{r}
#| label: fig-nfl-longevity
viz_nfl_roster_seasons_by_round(tar_read(picks_nfl), tar_read(nfl_rosters))
```

## Player dossier

`player_dossier()` resolves a player's scattered rows into one **interpreted**
record (decision 0019) rather than a raw extract that is easy to misread. Each
dossier shows: identity and physicals; college statistics grouped and labeled by
game **phase** (offense / defense / special teams); per-season coaching context;
name-guarded high-school recruiting; the NFL outcome (drafted or
undrafted-signed) with longevity; and a provenance / caveats block.

Below are the **#1 overall picks of the past five NFL drafts**, each rendered as
an interpreted dossier.

Notes:

- A bare `TD` in the long `player_stats` is ambiguous, so stats are relabelled
  by phase (e.g. "Passing touchdowns" vs "Defensive touchdowns").
- A blank coach cell is shown as "head coach not available" — the CFBD coaches
  feed is FBS head coaches only, so a non-FBS season (e.g. Cam Ward's 2021 FCS
  year at Incarnate Word) is flagged, not silently dropped.
- Recruiting is name-guarded (decision 0013); a genuinely unrated player stays
  unrated, never imputed.
- NFL longevity is **right-censored** for these recent picks — all are still
  active, so seasons/games are lower bounds.

```{r}
#| label: dossier-setup
#| include: false
# Read the shared inputs once; `dossier()` keys a player by CFBD athlete id.
ps <- tar_read(player_stats)
ros <- tar_read(roster)
co <- tar_read(coaches)
picks_nfl <- tar_read(picks_nfl)
rec <- tar_read(recruiting)
nrs <- tar_read(nfl_rosters)
nps <- tar_read(nfl_player_stats)
dossier <- function(id) {
  player_dossier(id, ps, ros, co, picks_nfl, rec, nrs, nps)
}
```

### Fernando Mendoza — 2026 #1, Indiana

```{r}
#| label: dossier-mendoza
#| comment: ""
dossier("4837248")
```

### Cam Ward — 2025 #1, Miami

```{r}
#| label: dossier-ward
#| comment: ""
dossier("4688380")
```

### Caleb Williams — 2024 #1, USC

```{r}
#| label: dossier-williams
#| comment: ""
dossier("4431611")
```

### Bryce Young — 2023 #1, Alabama

```{r}
#| label: dossier-young
#| comment: ""
dossier("4685720")
```

### Travon Walker — 2022 #1, Georgia

```{r}
#| label: dossier-walker
#| comment: ""
dossier("4426349")
```

## Branding — team logos

The `teams` dimension carries each program's logo and colors, joined by school
name via `link_team_meta()` and kept off the model path (display only, decision
0008). Two rendering paths: a `reactable` table with logo thumbnails and color
swatches, and a `ggiraph` bar filled with each team's brand color that shows its
logo on hover.

```{r}
#| label: tbl-team-logos
viz_team_logos(tar_read(teams))
```

```{r}
#| label: fig-drafts-by-team
viz_drafts_by_team(tar_read(model_table), tar_read(teams))
```

::: {.callout-tip}
These are templates, not findings — they demonstrate the wiring from a pipeline
target to an interactive widget. Real analyses will pair model inputs with model
results on the same page.
:::
