---
title: "Explore"
vignette: >
  %\VignetteIndexEntry{Explore}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(cfbstats)
library(dplyr)
library(targets)
root <- Sys.getenv("CFBSTATS_ROOT")
if (nzchar(root)) knitr::opts_knit$set(root.dir = root)
```

Interactive exemplars for the three chart families the project uses: **time
series**, **2D**, and **3D**. Each reads a pipeline target, so the input → chart
lineage is explicit. The default stack is R htmlwidgets (`plotly`, `ggiraph`);
Observable JS is reserved for bespoke explorables (a later addition).

```{r}
#| label: data
picks <- tar_read(picks) |>
  filter(year >= 2013, !is.na(height), !is.na(weight), !is.na(overall))
```

## Time series — drafts per year

```{r}
#| label: fig-timeseries
viz_drafts_by_season(tar_read(model_table))
```

## 2D — draft capital vs. physicals (hover for player)

Each point is a drafted player; hover for the name. Color encodes draft round.

```{r}
#| label: fig-2d
viz_scatter_2d(
  dplyr::slice_sample(picks, n = min(1500, nrow(picks))),
  x = weight, y = height, tooltip = name, color = round
)
```

## 3D — height × weight × overall pick

```{r}
#| label: fig-3d
viz_scatter_3d(picks, x = "height", y = "weight", z = "overall", color = "round")
```

## NFL outcomes — how long drafted players last

Each point is a drafted player bridged to nflverse (decision 0014), placed by
draft round against the number of distinct **seasons they spent on an NFL
roster** — the project's primary longevity measure. The **#1 overall picks** are
highlighted against the rest of the field; the grey bar marks each round's
median. Hover for the player. Recent draft classes are still active, so their
counts are **right-censored** lower bounds.

```{r}
#| label: fig-nfl-longevity
viz_nfl_roster_seasons_by_round(tar_read(picks_nfl), tar_read(nfl_rosters))
```

## Player dossier — an interpreted extract

A raw extract of a player's rows is easy to misread: the long `player_stats`
stores a bare `TD` under a `category`, so offensive, defensive, and
special-teams scores look alike; drafted players link cleanly to the NFL but
undrafted-yet-signed ones don't; and a blank coach cell could mean "no data" or
"non-FBS". `player_dossier()` resolves these into one **interpreted** record
(decision 0019): stats grouped and labeled by phase, per-season coaching with an
explicit "not available" where the FBS-only feed is blank, name-guarded
recruiting, a drafted/undrafted-signed NFL outcome, and a provenance block.

```{r}
#| label: dossier
#| comment: ""
# Cam Ward (CFBD athlete id 4688380): #1 overall in 2025, with a 2021 FCS
# season at Incarnate Word that surfaces the head-coach coverage gap.
player_dossier(
  "4688380",
  tar_read(player_stats),
  tar_read(roster),
  tar_read(coaches),
  tar_read(picks_nfl),
  tar_read(recruiting),
  tar_read(nfl_rosters),
  tar_read(nfl_player_stats)
)
```

## Branding — team logos

The `teams` dimension carries each program's logo and colors, joined by school
name via `link_team_meta()` and kept off the model path (display only, decision
0008). Two rendering paths: a `reactable` table with logo thumbnails and color
swatches, and a `ggiraph` bar filled with each team's brand color that shows its
logo on hover.

```{r}
#| label: tbl-team-logos
viz_team_logos(tar_read(teams))
```

```{r}
#| label: fig-drafts-by-team
viz_drafts_by_team(tar_read(model_table), tar_read(teams))
```

::: {.callout-tip}
These are templates, not findings — they demonstrate the wiring from a pipeline
target to an interactive widget. Real analyses will pair model inputs with model
results on the same page.
:::
