---
title: "Explore"
vignette: >
  %\VignetteIndexEntry{Explore}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(cfbstats)
library(dplyr)
library(targets)
root <- Sys.getenv("CFBSTATS_ROOT")
if (nzchar(root)) knitr::opts_knit$set(root.dir = root)
```

Interactive exemplars for the three chart families the project uses: **time
series**, **2D**, and **3D**. Each reads a pipeline target, so the input → chart
lineage is explicit. The default stack is R htmlwidgets (`plotly`, `ggiraph`);
Observable JS is reserved for bespoke explorables (a later addition).

```{r}
#| label: data
picks <- tar_read(picks) |>
  filter(year >= 2013, !is.na(height), !is.na(weight), !is.na(overall))
```

## Time series — drafts per year

```{r}
#| label: fig-timeseries
viz_drafts_by_season(tar_read(model_table))
```

## 2D — draft capital vs. physicals (hover for player)

Each point is a drafted player; hover for the name. Color encodes draft round.

```{r}
#| label: fig-2d
viz_scatter_2d(
  dplyr::slice_sample(picks, n = min(1500, nrow(picks))),
  x = weight, y = height, tooltip = name, color = round
)
```

## 3D — height × weight × overall pick

```{r}
#| label: fig-3d
viz_scatter_3d(picks, x = "height", y = "weight", z = "overall", color = "round")
```

::: {.callout-tip}
These are templates, not findings — they demonstrate the wiring from a pipeline
target to an interactive widget. Real analyses will pair model inputs with model
results on the same page.
:::
