---
title: "Pipeline"
vignette: >
  %\VignetteIndexEntry{Pipeline}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
library(cfbstats)
library(targets)
library(dplyr)
# Under altdoc, docs render in a temp dir; point the knit working directory at
# the project root so tar_read()/tar_mermaid() find the store, _targets.R, and R/.
root <- Sys.getenv("CFBSTATS_ROOT")
if (nzchar(root)) knitr::opts_knit$set(root.dir = root)
```

The processing pipeline is a [`targets`](https://books.ropensci.org/targets/)
graph (decision 0004) — the single source of truth for how raw data becomes
features, models, and results. The map below is **generated from the graph**, so
it can't drift from what the code actually does.

## Stages

```{mermaid}
flowchart LR
  ingest[ingest\nCFBD API] --> clean[clean\nstandardize + coerce keys]
  clean --> link[link\nplayer-season backbone + joins]
  link --> features[features\nlongitudinal change]
  features --> model[model\nexplain + predict]
  model --> report[report\nsite + viz]
  clean -.-> contracts[(data contracts)]
  clean -.-> audit[(audit log)]
  link -.-> audit
  features -.-> audit
```

## Live dependency graph

```{r}
#| label: dag
#| output: asis
cat("```{mermaid}", tar_mermaid(), "```", sep = "\n")
```

## What remains

Targets that are outdated (need to run) — a live view of what's left to build or
refresh.

```{r}
#| label: outdated
outdated <- tryCatch(tar_outdated(), error = function(e) character(0))
if (length(outdated)) {
  tibble::tibble(target = outdated)
} else {
  cat("All targets are up to date.")
}
```

::: {.callout-important}
## Known gaps
`/roster` is now ingested, keyed on an authoritative `season` (stamped from the
queried year — the endpoint's own `year` is the eligibility class). But roster
**weight is static per player**, so the decision-0003 between-season
`weight_delta` is not derivable and was dropped; we keep the static roster
`weight`. The `model` and `report` stages are wired placeholders.
:::
